Abstractions for Multi-Sorted Substitutions

### 2024-03-11 paper/paper.tex
Update.

  an implementation as an Agda framework featuring
  a reflection algorithm, representation independence for
  substitutions and typing contexts, heterogeneous equality between
  renamings and substitutions, and absence of postulated axioms.
  \begin{code}
fooo
  \end{code}

@@

  \subsection{A Structure for Multi-Sorted Syntax}
  \label{sec:syntax:structure}
  The regularity of the multi-sorted syntax makes it easy to define a
  structure for syntaxes of arbitrary object-languages, i.e.\ syntaxes
  with arbitrary amounts of syntactic categories and variables:
  \LibCode\KSyntax
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \ACode{\AField{`\_}} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FSyntaxInst

@@

  \subsection{Unifying Renamings \& Substitutions}
  \label{sec:maps:kits}
  To avoid the duplication between renamings and substitutions,
  McBride\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  introduced an abstraction called \emph{kits}.\footnote{
    While kits were originally formulated for intrinsically-typed languages,
    they nicely extend to our extrinsically-typed multi-sorted syntax, whose
    sort-indexing is a form of intrinsic typing.
  }

  % A \ACode{Kit} is a structure that captures the commonalities
  % between variables and terms:
  A \ACode{Kit} is a structure that allows to abstract over whether
  something is a term or a variable. The intention is to instantiate
  this structure exactly twice (once for variables and once for terms),
  and then write definitions, which are parameterized over a kit and
  consequently can be used for both variables and terms.
  \LibCode\KKit

@@

  A kit provides the following operations:
  \begin{itemize}
  \item 
    \ACode{\AField{id/`}} converts a variable \ACode{S \ACon{∋} s} into a
    \ACode{S \AField{∋/⊢} s}. For the variable kit, \ACode{S ∋/⊢ s} is instantiated to \ACode{S \ACon{∋} s}, so this
    operation is the identity. For the term kit, \ACode{S ∋/⊢ s} is
    instantiated to \ACode{S \AField{⊢} s}, so this operation is the
    variable constructor \ACode{\AField{`\_}}.
  \item 
    \ACode{\AField{`/id}} converts a \ACode{S ∋/⊢ s} into a term \ACode{S \AField{⊢} s}
    and is analogous to the \ACode{\AField{id/`}} operation.
  \item 
    \ACode{\AField{wk}} shifts the DeBruin-Indices in a \ACode{S ∋/⊢ s}.
    The new, unused variable \ACode{\ACon{zero}} can assume any sort \ACode{s'}.
    For variables, \ACode{\AField{wk}} is the successor \ACode{\ACon{suc}}.
    For terms, \ACode{\AField{wk}} means applying a shifting renaming to the term.
  \end{itemize}

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \ACode{\AField{`/`-is-`}} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \ACode{\AField{`\_}}.
  \item \ACode{\AField{`/id-injective}} and \ACode{\AField{id/`-injective}} state that
    \ACode{\AField{`/id}} and \ACode{\AField{id/`}} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \ACode{\AField{wk-id/`}} characterizes the behaviour of the
    \ACode{\AField{wk}} function by how it acts on variables.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

  Building on the operations and axioms of a \ACode{\ACon{Kit}}, we
  define the usual operations for renamings and substitutions:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  % \item 
  %   A map can be extended by a new variable or term:
  %   \LibCode\KExt
  % \item 
  %   A map can be shifted, as the \ACode{\ACon{Kit}} provides the
  %   necessary \ACode{\AField{wk}} operation:
  %   \LibCode\KWkm
  \item 
    A map can be lifted, which allows to
    move a map under a binder of sort \ACode{s}:
    % \LibCode\KLift
    \LibCode\KLiftAlt
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingleAlt
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeakenAlt
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
    \LibCode\KFunExt
  \item 
    There is an identity map:
    \LibCode\KId
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}
  We also specify notation 
  to explicitly pass a particular \ACode{\ACon{Kit}} to those definitions:
  % We also specify notation, which makes it less cumbersome
  % to explicitly pass a particular \ACode{\ACon{Kit}} to those definitions:
  \begin{itemize}
  \item
     we write \ACode{(S \ADef{∋/⊢[} K \ADef{]} s)} for the \ACode{(S ∋/⊢ s)}
     of some specific \ACode{\ACon{Kit}} \ACode{K}; and
   \item
     we write \ACode{(S \ADef{–[} K \ADef{]→} s)} for the \ACode{(S \ADef{→ₖ} s)}
     of some specific \ACode{\ACon{Kit}} \ACode{K}.
  \end{itemize}

@@

  First, we define the operation of applying a map to a term:
  \AppCode\FTraversalOp
  The interesting cases are those with variables and binders:
  \begin{itemize}
  \item 
    In the variable case \ACode{(\ACon{`} x) \ADef{⋯} ϕ}, we first apply the map
    \ACode{ϕ} to the variable \ACode{x}. If \ACode{ϕ} is a renaming,
    we get back a variable and need to apply the variable constructor
    \ACode{\ACon{`\_}} to it. If \ACode{ϕ} is a substitution,
    we get back a term that we can use directly.
    This is exactly what \ACode{\AField{`/id}} does.
  \item 
    In cases where the operation needs to go under a binder, like
    \ACode{(\ACon{λx} e) \ADef{⋯} ϕ}, we lift the map using \AField{\_↑\_} before
    we apply it to the subterm to account for the bound variable.
  \end{itemize}

  Then we prove that applying an identity map does not change the term:
  \AppCode\FTraversalId
  \AppCode\FTraversalIdProofInteresting
  The interesting cases are:
  \begin{itemize}
  \item
    the variable case \ACode{\ACon{`} x}, where we need to combine the
    \ACon{\AField{id/`}} from the identity map, with the \ACon{\AField{`/id}}
    from the traversal operation using \ACon{\AField{`/`-is-`}}.
  \item
    the lambda abstraction case \ACode{\ACon{λx} e}, where the
    traversal lifts the identity over its binder. Here
    \ACon{\AField{id↑-id}} is used to state that this lifted identity
    map is still the identity.
  \item
    all other cases are analogous and hence omitted.
  \end{itemize}

@@

  As we defined the \ACode{\ACon{Kit}} structure before the
  \ACode{\ACon{Traversal}} structure, the fields of
  \ACode{\ACon{Kit}} could not use the \ACode{\AField{\_⋯\_}} function in
  their types.

  This prevented us to define a useful lemma, which also needs to
  be instantiated separately for variables and terms: the counterpiece
  to \ACode{\AField{wk-id/`}}, which we call \ACode{\AField{wk-`/id}}.

@@

  The definitions reveal two interesting properties:
  \begin{enumerate}
  \item
    If we compose two maps \ACode{ϕ₁} and \ACode{ϕ₂}, then the
    resulting map is a renaming, iff both \ACode{ϕ₁} and \ACode{ϕ₂} are
    renamings.
    In other words:
    if \ACode{ϕ₁} is a \ACode{K₁}-map and \ACode{ϕ₂} is a \ACode{K₂}-map,
    then the result is a \ACode{(K₁ ⊔ K₂)}-map, where \ACode{⊔} refers to the
    lattice for \ACode{\{ \ADef{Kᵣ} , \ADef{Kₛ} \}}
    generated by \ACode{\ADef{Kᵣ} < \ADef{Kₛ}}.
  \item
    All four compositions first apply \ACode{ϕ₁} to \ACode{x},
    and then apply \ACode{ϕ₂} to the result.
    If \ACode{ϕ₁} is a renaming, then this result is another variable, but
    if \ACode{ϕ₁} is a substitution, then this result is a term.
  \end{enumerate}
  With the \ACode{\ACon{Kit}} abstraction, we can easily abstract over
  \ACode{ϕ₂} being a renaming or a substitution:
  \ExampleCode\KExTwoComps

>>>>

@@

  \subsection{A Structure for Multi-Sorted Syntax}
  \label{sec:syntax:structure}
  % The regularity of the multi-sorted syntax makes it easy to define a
  % structure for syntaxes of arbitrary object-languages, i.e.\ syntaxes
  % with arbitrary amounts of syntactic categories and variables:
  The regularity of the multi-sorted syntax makes it easy to define a
  structure for arbitrary syntaxes, i.e. with arbitrarily many syntactic categories and variable types:
  \LibCode\KSyntax
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \ACode{\AField{`\_}} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FSyntaxInst

@@

  \subsection{Unifying Renamings \& Substitutions}
  \label{sec:maps:kits}
  To avoid the duplication between renamings and substitutions,
  McBride\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  introduced the \emph{Kit} abstraction.%
  % \footnote{
  %   While kits were originally formulated for intrinsically-typed languages,
  %   they nicely extend to our extrinsically-typed multi-sorted syntax, whose
  %   sort-indexing is a form of intrinsic typing.
  % }
  % A \ACode{Kit} is a structure that captures the commonalities
  % between variables and terms:
  A \ACode{Kit} is a structure that allows to abstract over whether
  something is a term or a variable. The intention is to instantiate
  this structure exactly twice (once for variables and once for terms),
  and then write definitions, which are parameterized over a kit and
  consequently can be used for both variables and terms.
  \LibCode\KKit

@@

  A kit consists of the following components:
  \begin{itemize}
  \item 
    \ACode{\AField{id/`}} converts a variable \ACode{S \ACon{∋} s} into a
    \ACode{S \AField{∋/⊢} s}. For the variable kit, \ACode{S ∋/⊢ s} is instantiated to \ACode{S \ACon{∋} s}, so this
    operation is the identity. For the term kit, \ACode{S ∋/⊢ s} is
    instantiated to \ACode{S \AField{⊢} s}, so this operation is the
    variable constructor \ACode{\AField{`\_}}.
  \item 
    \ACode{\AField{`/id}} converts a \ACode{S ∋/⊢ s} into a term \ACode{S \AField{⊢} s}
    and is analogous to the \ACode{\AField{id/`}} operation.
  \item 
    \ACode{\AField{wk}} shifts the DeBruin-Indices in a \ACode{S ∋/⊢ s}.
    The new, unused variable \ACode{\ACon{zero}} can assume any sort \ACode{s'}.
    For variables, \ACode{\AField{wk}} is the successor \ACode{\ACon{suc}}.
    For terms, \ACode{\AField{wk}} means applying a shifting renaming to the term.
  \item \ACode{\AField{`/`-is-`}} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \ACode{\AField{`\_}}.
  \item \ACode{\AField{`/id-injective}} and \ACode{\AField{id/`-injective}} state that
    \ACode{\AField{`/id}} and \ACode{\AField{id/`}} are injective.
    % This property
    % follows easily, as both the identity function and the variable
    % constructor are injective.
  \item \ACode{\AField{wk-id/`}} characterizes the behaviour of the
    \ACode{\AField{wk}} function by how it acts on variables:
    injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

  \begin{figure}
    \label{fig:map-ops}
    \centering
    \begin{minipage}[t]{0.5\linewidth}
    \LibCode\KMap
    \LibCode\KAp
    \LibCode\KLiftAlt
    \LibCode\KSingleAlt
    \end{minipage}
    \begin{minipage}[t]{0.4\linewidth}
    \LibCode\KWeakenAlt
    \LibCode\KEq
    \LibCode\KFunExt
    \LibCode\KId
    \LibCode\KIdLift
    \end{minipage}
    \caption{Map Operations}
  \end{figure}
  Figure~\ref{fig:map-ops} shows the usual operations for renamings and substitutions
  generalized to maps.\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  The type \ACode{S₁ \ADef{→ₖ} S₂} unifies
    renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}.
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  The operation \ACode{ϕ \& x} applies a map to a variable.
  The operation \ACode{ϕ ↑ s} lifts a map under a binder of sort \ACode{s}.
  The operation \ACode{⦅ x/t ⦆} constructs a singleton map that replaces
  \ACode{\ACon{zero}} with \ACode{x/t} and decreases all other variables by one.
  The \ACode{\AField{weaken s}} map increases all variables by one.
  The \ACode{ϕ₁ $\sim$ ϕ₂} type expresses extensional equality of maps.
    For simplicity, we postulate functional extensionality \ACode{\~-ext}.\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
  There is an identity map \ACode{id}.
  The lemma \ACode{id↑-id} states that a lifted identity map is again an identity map.

  To make it easier to talk about a specific kit, we introduce the following notations:
  % We also specify notation 
  % to explicitly pass a particular \ACode{\ACon{Kit}} to those definitions:
  % We also specify notation, which makes it less cumbersome
  % to explicitly pass a particular \ACode{\ACon{Kit}} to those definitions:
  \begin{itemize}
  \item
     we write \ACode{(S \ADef{∋/⊢[} K \ADef{]} s)} for the \ACode{(S ∋/⊢ s)}
     of some specific \ACode{\ACon{Kit}} \ACode{K}; and
   \item
     we write \ACode{(S₁ \ADef{–[} K \ADef{]→} S₂)} for the \ACode{(S₁ \ADef{→ₖ} S₂)}
     of some specific \ACode{\ACon{Kit}} \ACode{K}.
  \end{itemize}

@@

  First, we define the operation of applying a map to a term:
  \AppCode\FTraversalOp
  The interesting cases are those with variables and binders:
  \begin{itemize}
  \item 
    In the variable case \ACode{(\ACon{`} x) \ADef{⋯} ϕ}, we first apply the map
    \ACode{ϕ} to the variable \ACode{x}. If \ACode{ϕ} is a renaming,
    we get back a variable and need to apply the variable constructor
    \ACode{\ACon{`\_}}. If \ACode{ϕ} is a substitution,
    we get back a term that we can use directly.
    This is exactly what \ACode{\AField{`/id}} does.
  \item 
    In cases where the operation needs to go under a binder, like
    \ACode{(\ACon{λx} e) \ADef{⋯} ϕ}, we lift the map using \AField{\_↑\_} before
    we apply it to the subterm to account for the bound variable.
  \end{itemize}

  Then we prove that applying an identity map does not change the term:
  \AppCode\FTraversalId
  \AppCode\FTraversalIdProofInteresting
  We only display and discuss the interesting cases:
  \begin{itemize}
  \item
    in the variable case \ACode{\ACon{`} x}, the
    \ACon{\AField{id/`}} from the identity map meets the \ACon{\AField{`/id}}
    from the traversal operation, so we use \ACon{\AField{`/`-is-`}}.
  \item
    in the lambda abstraction case \ACode{\ACon{λx} e}, the
    traversal lifts the identity under its binder. Here
    we need to use \ACon{\AField{id↑-id}} to show that a lifted identity
    map is again an identity map.
  \end{itemize}

@@

  As we defined the \ACode{\ACon{Kit}} structure before the
  \ACode{\ACon{Traversal}} structure, the fields of
  \ACode{\ACon{Kit}} could not use the \ACode{\AField{\_⋯\_}} function in
  their types.
  This prevented us to define a useful lemma, which also needs to
  be instantiated separately for variables and terms: the counterpiece
  to \ACode{\AField{wk-id/`}}, which we call \ACode{\AField{wk-`/id}}.

@@

  The definitions reveal two interesting properties:
  \begin{enumerate}
  \item
    If we compose two maps \ACode{ϕ₁} and \ACode{ϕ₂}, then the
    resulting map is a renaming, iff both \ACode{ϕ₁} and \ACode{ϕ₂} are
    renamings.
    In other words:
    if \ACode{ϕ₁} is a \ACode{K₁}-map and \ACode{ϕ₂} is a \ACode{K₂}-map,
    then the result is a \ACode{(K₁ ⊔ K₂)}-map, where \ACode{⊔} refers to the
    lattice for \ACode{\{ \ADef{Kᵣ} , \ADef{Kₛ} \}}
    generated by \ACode{\ADef{Kᵣ} < \ADef{Kₛ}}.
  \item
    All four compositions first apply \ACode{ϕ₁} to \ACode{x},
    and then apply \ACode{ϕ₂} to the result.
    If \ACode{ϕ₁} is a renaming, then this result is another variable, but
    if \ACode{ϕ₁} is a substitution, then this result is a term.
  \end{enumerate}
  With the \ACode{\ACon{Kit}} abstraction, we can easily abstract over
  \ACode{ϕ₂} being a renaming or a substitution:

  \noindent
  \begin{minipage}[t]{0.60\linewidth}
  \ExampleCode\KExTwoCompsI
  \end{minipage}
  \begin{minipage}[t]{0.40\linewidth}
  \ExampleCode\KExTwoCompsII
  \end{minipage}
